---
title: "Lab_7: K-Means Clustering"
author: "Ryan Waterman"
date: "2025-10-24"
output: html_document
---
# K-Means Clustering

In this lab, we use the ‘NbClust’ and ‘kmeans’ R functions to implement k-means clustering on the lodgepole
pine data. We first load the R libraries we will be using. The ‘NbClust’ function is available in the ‘NbClust’
R library and the ‘kmeans’ function is available in base R. The ‘tidymodels’ and ‘dplyr’ R libraries are also
loaded, since we will be using the ‘intitial_split’ function (from ‘tidymodels’) to partition the data into
training and test sets, as well as the ‘select’ function (from ‘dplyr’) to select specific columns.

```{r}
library(tidymodels)
library(dplyr)
library(NbClust)
```

*K-Means Clustering*

Let’s see if the observations can be clustered based on their horizontal distance to important landmarks (horizontal_distance_to_hydrology, horizontal_distance_to_roadways, and horizontal_distance_to_fire_points). In practice, we can cluster based on any features we think are relevant; this is where the subjective part of unsupervised learning comes into play, since our analysis is often guided by subject matter knowledge and expertise.

Before we implement k-means clustering, we center and scale the features in the training and test sets, which is critical for k-means clustering for the reasons discussed during the live session. We also only keep the features that will be included in the clustering analysis: horizontal_distance_to_hydrology, horizontal_distance_to_roadways, and horizontal_distance_to_fire_points.

```{r}
data <- read.csv("lodgepole_data.csv")
split <- initial_split(data, prop = 0.75, strata = lodgepole_pine)
training_set <- training(split) %>% select(horizontal_distance_to_hydrology,
horizontal_distance_to_roadways,
horizontal_distance_to_fire_points)
test_set <- testing(split) %>% select(horizontal_distance_to_hydrology,
horizontal_distance_to_roadways,
horizontal_distance_to_fire_points)
# center and scale the features
test_set <- scale(test_set,
center = apply(training_set, 2, mean),
scale = apply(training_set, 2, sd))
training_set <- scale(training_set)

```

Now that the features have been centered and scaled, we implement k-means in R using the ‘NbClust’ and
‘k-means’ functions. We first use the ‘NbClust’ function to determine if there is an optimal number of clusters
according to the 26 criteria incorporated into this function. We use a random sample of 1000 rows of the
original training set, since the ‘NbClust’ function can take a long time to run on large datasets. We set the
seed for random number generation, so the exact same results will be obtained when run on your machine
(otherwise, the results may be different, since the sample of 1000 rows may be different).

```{r}
set.seed(123)
nc <- NbClust(training_set[sample(nrow(training_set), 1000), ],
              min.nc = 2, max.nc = 10, method = "kmeans")
```

```{r}
km_clusters <- kmeans(training_set, centers = 4)
```

After calculating the clusters, we can append the cluster labels to the training set. Now, we can use the
cluster labels as a categorical feature in a supervised learning model (such as a dense feed-forward neural
network). As with any other categorical feature, we need to encode the cluster label (such as using one-hot
encoding) before applying a supervised learning model.

```{r}
cluster_number <- data.frame(cluster_number = km_clusters$cluster)
training_set_with_cluster_num <- cbind(training_set, cluster_number)
head(training_set_with_cluster_num)
```

In this lab, we used the ‘NbClust’ and ‘kmeans’ functions to implement k-means clustering. However, several
different methods for implementing clustering exist and are also implemented in R. Look into the different
methods, as well as other arguments, that are implemented in the ‘NbClust’ function discussed above.

*Exercises*

1) Why is it good practice to center and scale before applying k-means clustering?

*Centering is good practice because it shifts the mean value of each feature to 0, which doesn't directly affect the K-Means clustering algorithm, but is necessary for scaling. Scaling is far more critical for K-Means clustering, as it ensures that larger features cannot have disproportionately large clusters.*

2) Print the cluster sizes and centers to the R console. Include a screenshot of the output.
```{r}
km_clusters$size
km_clusters$centers
```

3) Use the aggregate() function to calculate the mean of each variable within each cluster. Include a
screenshot of the output.

```{r}
aggregate(. ~ cluster_number, 
          data = training_set_with_cluster_num, 
          FUN = mean)
```

4) Assign the observations in test_set to the clusters found using the sample of the observations in
training_set above. Include a screenshot of the output. Hint: a function in the ‘clue’ R library can be
used for this.

```{r}
library(clue)

test_cluster_assignments <- cl_predict(km_clusters, test_set)
head(test_cluster_assignments)

test_set_with_cluster_num <- cbind(as.data.frame(test_set), cluster_number = test_cluster_assignments)
head(test_set_with_cluster_num)
```



